---
lang: en
format: 
  pdf:
    linestretch: 1.25
    fig-format: pdf
    mainfont: Roboto
    include-in-header: ../styles.tex
    keep-tex: true
fig-height: 2.8
fig-width: 7.9
crossref:
  fig-title: Fig.
  fig-prefix: Fig.
knitr:
  opts_chunk: 
    dev: cairo_pdf
execute:
  echo: false
  warning: false
bibliography: ../refs/references.bib
csl: ../refs/science-without-titles.csl
appendix-style: none
---

{{< include ../_init.qmd >}}

```{r}
#| label: stats

df_stats <- trs_global$get_statistical_details(
  vars = c("dtr", "ts", "past_dmat"),
  ELEV_SPAN_DEFAULT,
  EXCL_DEFAULT,
  std_from = "top"
)
global_stats <- partial(get_regression_stats, df_stats)

df_slope_summary <- trs_local$get_slope_summary_()
slope_perc <- partial(get_slope_perc, df_slope_summary)
```

# Diurnal temperature variation drives elevational range sizes globally

**Authors:**

```{r}
#| results: asis
authors$get_author_list(format = "^a^c", FALSE) %>% cat(sep = ", ")
```

**Affiliations:**

```{r}
#| results: asis
authors$get_affiliations() %>% cat(sep = "\n\n")
```

\*Corresponding author. Email: `r authors$get_corresponding_emails()`

**Abstract:**

A prominent hypothesis in ecology is that larger species ranges are found in more variable climates because species develop broader environmental tolerances, predicting a positive range size-temperature variability relationship. However, this overlooks the extreme temperatures that variable climates impose on species, with upper or lower thermal limits more likely to be exceeded. Accordingly, we propose the 'temperature range squeeze' hypothesis, predicting a negative range size-temperature variability relationship. We test these contrasting predictions by relating `r data_info$n_data` elevational range sizes of vascular plants in `r data_info$n_locations` mountains to short- and long-term temperature variation. Consistent with our novel hypothesis, we find that species' range size is negatively correlated with diurnal temperature range. Accurate predictions of short-term temperature variation will become increasingly important for extinction risk assessment in the future.

**One-Sentence Summary:**

Decreasing elevational range sizes of plants with greater diurnal temperature range supports a novel biodiversity hypothesis and indicates elevated extinction risk of continental species.

**Main Text:**

There is remarkable variation in species' range sizes, from endemic species confined to single mountaintops to cosmopolitan species occurring in most habitats around the world [@brown1996]. Although variation in species distributions has been highlighted since the earliest days of ecology and biogeography [@vonhumboldt1805], the question of how range size dynamics relate to climate and geography has remained controversial [@gaston1999; @colwell2000; @mccain2009; @klinges2021; @macek2021]. Progress on this question is essential for better understanding of ecological and evolutionary processes, broad-scale diversity patterns [@stevens1992; @rahbek2019] and species' extinction risks in the context of climate change [@urban2015; @staude2020].

Temporal thermal variation has often been used to explain variance in species' range sizes [e.g. @mccain2009; @morin2011; @pintor2015; @hernández-rojas2020; @liang2021; @maccagni2022]. In his seminal contribution "Why mountain passes are higher in the tropics", Janzen [@janzen1967] suggested that species experiencing large temperature fluctuations evolve broader thermal tolerances than species inhabiting ecosystems with relatively constant temperatures ([@fig-conceptual]A). Stevens [@stevens1992; @stevens1989] used this argument to explain observations that species' geographical ranges increase towards high latitudes or elevations [known as Rapoport Rule, @rapoport1982]. He hypothesized that the broader thermal tolerance of species inhabiting thermally variable environments allows these species to survive in a broader range of latitudes and elevations ([@fig-conceptual]B; hereafter "Stevens' hypothesis").

While a direct link between species' thermal tolerances, temperature variability and occupied geographic ranges is appealing, Stevens' hypothesis misses a key point. Stevens' hypothesis neglects lethal temperatures as a limit for any given species' range size and that temperature variability differs between locations. For example, a location with mean temperature of +5°C on a mountain with high thermal variability may experience temperatures from -5°C to +15°C during the year or day ([@fig-conceptual]C). At that location, a species with a lethal lower temperature limit of 0°C and upper limit of 20°C will thus experience lethally cold temperatures frequently. Lower down, a similarly variable location with mean +20°C will vary from +10°C to +30°C annually or daily. There, that same species will experience lethally high temperatures. The range of elevations in which the species does not experience lethal temperatures in this thermally variable mountain is smaller than in an equivalent, more thermally constant mountain ([@fig-conceptual]C). Thus, greater temperature variability is a mechanism that will tend to reduce species' elevational ranges. This counteracts Stevens' expectation ([@fig-conceptual]B) that larger-ranged species are expected to be found in climatically more variable mountain systems. It leads us to propose that the opposite pattern may instead be found ([@fig-conceptual]D). We call this the "temperature range squeeze" hypothesis. This predicted pattern may be dampened by the larger thermal tolerances of species in more variable environments proposed by Janzen [@janzen1967] or by avoidance strategies such as dormancy, but would not be reversed unless the increase in tolerance over-compensates for the range squeeze.

```{r}
#| label: fig-conceptual
#| fig-height: 4
#| fig-width: 9.3
#| fig-cap: "**Illustration of Stevens' hypothesis (A-B) and the extreme temperature sensitivity hypothesis (C-D).** Based on the assumption that species have larger thermal tolerances in climatically variable habitats than in less climatically variable ones (**A**), Stevens predicted a positive relationship between species’ range sizes and temperature variability (**B**). Panel **C** represents the spatio-temporal temperature variation in mountains with different levels of temperature fluctuation. For simplicity, we used a constant lapse rate of 0.5°C. Shaded areas represent tolerable (blue) and lethal (red) temperatures for a species' thermal tolerance spanning from 0°C to 20°C. The widths of the shaded areas represent temperature variability over time (∆T) while the spatial variation in temperatures is described by the vertical axes. Thick blue and red lines indicate the 0°C and 20°C isotherms, respectively. The suitable habitat, which corresponds to species' maximum theoretical elevational range in the absence of avoidance strategies such as dormancy, is defined by elevations in which the species does not encounter a limiting temperature (i.e. below 0°C or above 20°C) at any time of the given time scale. The elevation span of the suitable habitat is expected to shrink as thermal variability increases (**C**, compare mountains 1 and 2). Thus, a negative relationship between species' range sizes and thermal variation is expected (**D**)."

theme_update(
  axis.title = element_text(size = 12),
  axis.text = element_text(size = 12)
)

####  Stevens hypothesis  ####
  {
    SH_base <- ggplot() + geom_segment(aes(x = 0, xend = 1, y = 0, yend = 1))
    
    # · Thermal tolerance-thermal variation rel. ----
    
    A <- SH_base +
      tag("A", "topleft", size = 14 / .pt) +
      lims(x = c(0, 1), y = c(0, 1)) +
      labs(
        x = "Temperature variation",
        y = "Thermal tolerance"
      ) +
      theme(
        axis.ticks = element_blank(),
        axis.text = element_blank()
      )
    
    # · Species' range-thermal variation rel. ----
    
    B <- SH_base +
      tag("B", "topleft", size = 14 / .pt) +
      ylab("Species' range size") +
      scale_x_continuous(
        breaks = .5,
        labels = "Temperature variation"
      ) +
      theme(
        axis.ticks = element_blank(),
        axis.text.y = element_blank(),
        axis.title.x = element_blank()
      )
  }

####  TRS hypothesis  ####
  {
    # · Conceptual fig. ----
    
    C <- make_conceptual_fig("trs", label_size = 11 / .pt) + 
      theme(axis.text.x = element_text(size = 12)) +
      tag("C", "topleft", size = 14 / .pt)
    
    # · Species' range-thermal variation rel. ----
    
    D <- ggplot() +
      geom_segment(aes(x = 0, xend = 1, y = 1, yend = 0)) +
      tag("D", "topright", size = 14 / .pt) +
      lims(x = c(0, 1), y = c(0, 1)) +
      labs(
        x = "Temperature variation",
        y = "Species' range size"
      ) +
      theme(
        axis.ticks = element_blank(),
        axis.text = element_blank()
      )
  }
  
####  Make fig.  ####
  {
    layout <- "
      A#CC#D
      B#CC##
      "
    patchwork::wrap_plots(A, B, C, D, design = layout) +
      patchwork::plot_layout(widths = rep(c(1, .1, 1), 2))
  }

theme_update(
  axis.title = element_text(size = 11),
  axis.text = element_text(size = 11)
)
```

Here we assess the contrasting predictions of how range size scales with temperature variability (herein we use this term for temporal thermal variability) using a global dataset of more than `r data_info$n_data` elevational range size estimates for vascular plants in `r data_info$n_locations` mountains (`r data_info$n_continents` continental and `r data_info$n_islands` island mountains). Each range size estimate is the difference between the maximum and minimum elevations of occurrence records in a given mountain. 'Mountains' are defined as mountainous areas, regions or countries, mountain ranges or volcanoes (see Methods and Table S1 for details) with an elevation span ≥ `r ELEV_SPANS[1]` m (Fig. S1). We used diurnal and seasonal temperature variations (mean diurnal range of temperatures averaged over one year and standard deviation of the monthly mean temperatures, respectively) as the main predictors of species' range sizes, as originally suggested by Stevens [@stevens1989]. Because most plants are perennial and experience similar diurnal and seasonal temperatures during their lifespan [within-generation variation, @gilchrist1995], we also tested the influence of temperature variation on species' range sizes among generations from 0 to 1980 AD (∆MAT~0-1980~). Most previous studies tested the relationship of plant species' elevational ranges with thermal variation along single mountain gradients, which makes the decoupling of climatic and geographic factors difficult. These limitations are best addressed with global-scale analyses and multiple elevational gradients: our core analysis. In addition, for direct comparability with Stevens' original hypothesis and previous studies, we tested the response of species' elevational ranges along each elevational gradient.

We asked how diurnal temperature range, temperature seasonality, and ∆MAT~0-1980~ predict the elevational ranges of vascular plants averaged within mountains across elevational gradients (global-scale analyses). Diverging conclusions in previous studies that investigated the relationship between temperature variability and species' range sizes have been attributed to methodological issues, such as sampling effort [@mccain2010], geometric constraints [@colwell2000; @macek2021] and analytical methods [@chan2016; @field2018]. To minimize the influence of these factors, we standardized the length of each elevational gradient to `r ELEV_SPAN_DEFAULT` m and discarded species found exclusively in the upper and lower `r EXCL_DEFAULT` m (exclusion zones; see Supplementary materials for details and justifications). We ran parallel analyses using standardized gradient lengths of `r enumerate(ELEV_SPANS[1:2])` m and exclusion zones of `r enumerate(EXCLS[c(1, 3)])` m, which did not influence the main results (see Supplementary materials for results using different data constraints).

Species' elevational range declined with all measures of temperature variability used in the study (@fig-regressions_main), disproving Stevens' hypothesis and supporting the temperature range squeeze hypothesis. In addition, models using temperature variability alone performed better than models including interactions with annual precipitation and mean annual temperature (Table S2), suggesting that additional climatic variables did not increase the predictive ability of the model. A higher probability for species to encounter limiting temperatures sooner as they move away from their optimum habitat likely explains the reduced elevational range sizes observed in thermally variable mountains.

```{r}
#| label: fig-regressions_main
#| fig-cap: !expr glue("**Relationships between thermal variability and mean species' elevational ranges.** Diurnal temperature range (**A**), temperature seasonality (**B**) and the variation of mean annual temperature from 0 to 1980 AD (**C**). Points represent the estimated mean elevational ranges with their respective standard error in each of the {data_info$n_locations_global} standardized elevational gradients with length ≥ {ELEV_SPAN_DEFAULT} m. Thick lines are the posterior mean calculated from {N_DRAWS} random draws sampled from the {CI_DEFAULT * 100}% credible interval (thin blue lines). Colored dots indicate island (white) and continental (gray) mountains.")

trs_global$regressions(
  vars = c("dtr", "ts", "past_dmat"),
  elev_span = ELEV_SPAN_DEFAULT,
  excl_zone = EXCL_DEFAULT,
  std_from = "top",
  labels = fig_labels
) |> 
  label_facets(LETTERS, x = Inf, hjust = 1)
```

Among thermal predictors, diurnal temperature range was the most predictive (lowest WAIC and LOO; Table S2) and had the strongest and least uncertain relationship to species' range size ($P(\beta < 0) = `r global_stats("dtr", "p_beta")`$, $R^2:`r global_stats("dtr", "r2")`$), followed by temperature seasonality ($P(\beta < 0) ≈ `r global_stats("ts", "p_beta")`$, $R^2:`r global_stats("ts", "r2")`$) and ∆MAT~0-1980~ ($P(\beta < 0) ≈ `r global_stats("past_dmat", "p_beta")`$, $R^2:`r global_stats("past_dmat", "r2")`$; @fig-regressions_main, Fig. S6 & Table S3). The strong relationship of species' range size to diurnal temperature range suggests that short-term temperature variation is more important in determining species range sizes than previously thought. Longer-term variables such as temperature seasonality might be less relevant than expected due to the ability of plant species inhabiting seasonal climates to avoid long periods of unfavorable temperatures via dormancy. As a consequence, the relative range of temperatures effectively experienced by extratropical species may be less than the estimated seasonal variability. The narrowing of sessile species' elevational ranges may be exacerbated in mountains with high diurnal temperature variation because they cannot escape extreme temperatures over such short time-scales. A decline in species' elevational ranges in response to high diurnal temperature range has also been reported in terrestrial vertebrates [@chan2016], suggesting a general pattern. Chan et al. [@chan2016] used a simulation by Gilchrist [@gilchrist1995] to explain the negative relationship via evolutionary selection by diurnal temperature fluctuation for narrow thermal tolerance. Gilchrist's simulation assumes survival of all individuals under any circumstances and variability in range size is merely linked to how reproductive success is related to performance breadth [@gilchrist1995]. Predictions are thus only realistic if species survive unsuitable temperatures (e.g. via dormant stages). If this assumption is not met, species with small elevational ranges need broad thermal tolerance in order to survive the extremes of short-term temperature fluctuation. Gilchrist makes explicit assumptions of the performance breadth of a species based on the interactions between among- and within-generation variation. Exploring our data further to assess the influence of these interactions did not show support for Gilchrist's hypothesis (Fig. S8-10).

```{r}
#| label: fig-regressions_land_type
#| fig-cap: !expr glue("**Relationships between thermal variability and mean species' elevational ranges in island and continental mountains.** Diurnal temperature range (**A**), temperature seasonality (**B**) and the variation of mean annual temperature from 0 to 1980 AD (**C**). Points represent the estimated mean elevational ranges with their respective standard error in each of the {data_info$n_locations_global} standardized elevational gradients with length ≥ {ELEV_SPAN_DEFAULT} m. Colors indicate island (white dots, blue regressions) and continental (gray dots, red regressions) mountains.")

trs_global$regressions(
  vars = "(dtr|ts|dmat)-land_type",
  elev_span = ELEV_SPAN_DEFAULT,
  excl_zone = EXCL_DEFAULT,
  by_land_type = TRUE,
  labels = fig_labels
) |> 
  label_facets(LETTERS, x = Inf, hjust = 1)
```

Island diversity differs from mainlands as a result of isolation-driven immigration, extinction and speciation processes [@flantua2020], which could influence species' range size differently. Comparing species' range sizes in island and continental mountains revealed broader mean species' range sizes on islands (@fig-regressions_main). However, diurnal temperature range was the only variable to account for variation in species' range sizes between island and continental mountains. Island mountains were characterized by both lower diurnal temperature range and broader mean elevational range sizes than continental mountains (@fig-regressions_land_type). Lower diurnal temperature variation in island mountains likely results from a buffering effect of the ocean on air temperature that would make the climate less temporally variable [@shen2021; @weigelt2013]. The strength and continuity in the diurnal temperature range-species' range size relationship in island and continental mountains ([@fig-regressions_land_type]A) suggest that no additional island-specific factors (e.g. lower competition) are needed to explain the range-size pattern and that the temperature range squeeze hypothesis offers a unifying explanation for elevational range sizes on continents and islands at a global scale.

We also assessed the relationship between temperature variation and species' elevational range within each individual mountain (local-scale analyses) using diurnal temperature range and temperature seasonality. We did not test the influence ∆MAT~0-1980~ on species' elevational range for the local-scale analyses because of the low spatial resolution of the past climate data. Because each mountain gradient was tested independently, we did not standardize elevational gradients. For the same reasons mentioned in the global-scale analyses, we ran parallel analyses using exclusion zones of `r enumerate(EXCLS[c(1, 3)])` m (see Supplementary materials for results using different data constraints).

Results from the local-scale analyses were less conclusive than the analyses at a global scale, with about as many mountains featuring negative as positive influence of diurnal temperature range (`r slope_perc("dtr", "+")`% positive vs `r slope_perc("dtr", "-")`% negative, including `r slope_perc("dtr", "+", "low")`% positive vs `r slope_perc("dtr", "-", "low")`% negative with low uncertainties) or temperature seasonality (`r slope_perc("ts", "+")`% positive vs `r slope_perc("ts", "-")`% negative, including `r slope_perc("ts", "+", "low")`% positive vs `r slope_perc("ts", "-", "low")`% negative with low uncertainties) on species' elevational ranges (@fig-histogram_slopes, Table S4). The lack of a clear trend in the direction of the estimates for all elevational gradients suggests that, at a local scale, temperature variability might not be the dominant driver of species' elevational ranges, or has different effects depending on plant community composition. Our results thus potentially explain ambiguous conclusions from previous studies on the role of thermal variation in shaping species' elevational ranges along individual mountain gradients. The small variation in our data on diurnal and seasonal temperature range within single mountain gradients compared to the variations across mountains could explain the weak influence of thermal variability in driving species' elevational ranges locally (@fig-histogram_slopes, Fig. S11). Additionally, species' ranges are likely to be co-determined by multiple factors that are mountain-dependent, such as land use, soil type, water availability, dispersal abilities, ecological history, microgeography or interspecific interaction [@klinges2021; @mccain2010; @sexton2009; @sheth2020; @chauvier2021; @lynn2019; @whitman2021].

```{r}
#| label: fig-histogram_slopes
#| fig-height: 2.3
#| fig-width: 4.6
#| fig-cap: !expr glue("**Results from the local-scale analyses.** Responses of species' elevational ranges to diurnal temperature range (**A**) and temperature seasonality (**B**) within each of the {data_info$n_locations} mountains used in the study. Histograms are of mean slope terms divided by their standard deviation from the {CI_DEFAULT * 100}% credible interval for the local-scale analyses. Colors indicate whether the {CI_DEFAULT * 100}% credible interval includes 0 (light blue) or not (dark blue). Dashed lines indicate 0.")


slope_hist <- trs_local$slope_histograms(excl_zone = EXCL_DEFAULT) +
  scale_y_continuous(breaks = seq(0, 12, 4)) +
  theme(
    strip.text = element_blank(),
    axis.title = element_text(size = 8.5),
    axis.text = element_text(size = 7.5)
  )
label_facets(slope_hist, LETTERS, size = 8.5 / .pt)
```

The ratio of positive and negative relationships between species' range sizes and temperature variation remained essentially the same, independent of the size of the exclusion zone (Fig. S12, Table S4). However, the number of slopes with low uncertainties decreased drastically with increasing exclusion zone sizes (Fig. S12, Table S4), suggesting that species found exclusively at the domain boundaries strongly influence the strength of the pattern within mountains and that the narrowing of species' range sizes at domain boundaries greatly influence the probability of finding statistically significant relationships. Species exclusively inhabiting lower and higher elevations had less influence in the global-scale analyses, where results remain consistent, independent of the size of the exclusion zone.

In summary, our findings support the novel temperature range squeeze hypothesis ([@fig-conceptual]C-D) that predicts a decline in species' range sizes in climatically variable habitats, thereby disproving Stevens' hypothesis [@stevens1989] ([@fig-conceptual]B). Diurnal temperature variation is important in driving the elevation ranges of vascular plants globally, and helps to account for range-size differences between continents and islands. Local (within-mountain) variation in species' elevational ranges, however, is largely decoupled from temperature variability. These advances in our understanding of the drivers of plant species distributions have broad implications for conservation biology. Our findings challenge the current understanding of how global change threatens small-ranged tropical species [@laurance2011]. Extinction risk may be higher in continental mountains where species are more likely to have smaller elevational ranges due to higher thermal variability than species inhabiting mountain systems exposed to oceanic climates. Short-term temperature variation may also become increasingly important in driving local extinction risks.

**References and Notes**

::: {#refs}
:::

**Acknowledgements:**

We are grateful to Monika Kriechbaum and Empar Carrillo Ortuño for providing us with species' authorities for accurate taxa standardization of the Mustang and Espot y Boí data, respectively. We thank Kirsten O'Sullivan, Chi-Hua Chang and the Forest Management Laboratory at the National Pingtung University of Science and Technology for translating, processing and putting together the Taiwan data. We thank Cathy Jenks for putting together the Drakensberg, Jamaica, Jaya and Hengduan datasets as well as Marten Winter for putting together the Afghanistan dataset. We thank Richard J. Telford for providing advice on the R code and figures as well as Camila Pacheco-Riaño for advice on GIS analyses.

**Funding:**

S.G.A.F. acknowledges support from Trond Mohn Stiftelse (TMS) and University of Bergen for the startup grant 'TMS2022STG03', and from the European Research Council (ERC) under the European Union's Horizon 2020 research and innovation program (grant agreement no. 741413) to H.J.B. Birks.

**Author contributions:**

```{r}
#| results: asis
contributions <- authors$get_contributions()
cat(contributions[c(1, 2, 6, 5, 4, 10, 11, 7, 3, 9, 8)], sep = "\n\n")
```

All authors contributed to the review & editing of the original draft.

**Competing interests:**

The authors declare no competing interests.

**Data and materials availability:**

Data will be publicly available after acceptance.

**Supplementary Materials**

Materials and Methods

Supplementary Text

Figs. S1 to S13

Tables S1 to S4

Data S1 to S2

References (`r ref_range("manuscript/refs/references.bib", from = "supplementary", sep = "--")`)
